A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950, including the Stensele station and the inland

### 2022-02-05 getdata2.Rmd
add biometric and new temperatures for Inland

We have two stations, *Umeå* and *Bjuröklubb*.

```{r gettemp}
source("R/tempdat.R")
temp_start <- 1894
temp_end <- 1951
temp_ivl <- c(temp_start, temp_end)
##
umetemp <- tempdat("ume", temp_ivl)
bjurtemp <- tempdat("bjur", temp_ivl)
##stentemp <- tempdat("~/Forskning/Data/sten_temp.csv", c(1899, 1952))
saveRDS(umetemp, file = "data/umetemp.rds")
saveRDS(bjurtemp, file = "data/bjurtemp.rds")
```

>>>>

We have three stations, *Umeå*, *Bjuröklubb*, and *Stensele*.

```{r gettemp}
source("R/tempdat.R")
temp_start <- 1894
temp_end <- 1951
temp_ivl <- c(temp_start, temp_end)
##
umetemp <- tempdat("ume", temp_ivl)
bjurtemp <- tempdat("bjur", temp_ivl)

##Changed 5 feb 2022:
##stentemp <- tempdat("sten", temp_ivl) # Added 27 Jan 2022
source("R/normal.R")
stentemp <- normal(temp_ivl)
## End change

saveRDS(umetemp, file = "data/umetemp.rds")
saveRDS(bjurtemp, file = "data/bjurtemp.rds")
saveRDS(stentemp, file = "data/stentemp.rds")  ## Added 27 Jan2022
```

### 2022-02-05 getdata3.Rmd
add biometric and new temperatures for Inland

2.    (January 27. 2022) I found that teperature reports from *Stensele* are very similar
to those from the very similar *Umeå* and *Skellefteå*. Conclusion: The Skellefteå inland is now included.

>>>>

2.    (January 27. 2022) I found that temperature reports from *Stensele* are very similar
to those from the very similar *Umeå* and *Skellefteå*.

3.    (February 5, 202). Conclusion: The Skellefteå inland is now included. But with 
summer temperatures from *Bjuröklubb*.

### 2022-02-05 paper3.Rmd
add biometric and new temperatures for Inland

Figure \@ref(fig:map) shows the study area within Sweden, with the weather stations 
marked. The map is taken from the paper by @junkka.

```{r map, fig.height = 5, fig.cap = "Umeå, Skellefteå (orange) and its inland (green)."}
include_graphics("myfigs/skum.png")
```

@@

```{r neoanalw, results = 'asis'}
##library(survival)
winter <- neo[neo$season %in% c("winter", "spring"), ]
winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = winter)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanalw", keep = 3)
```

@@

```{r neoanals, results = 'asis'}
##library(survival)
summer <- neo[neo$season %in% c("summer", "fall"), ]
summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = summer)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanals", keep = 2)
```

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = winter, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = 4)
```

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = summer, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = 4)
```

@@

A modern version of the Bourgeois-Pichat biometric modelling [@bp51a; @bp51b] is given here,
*without* the "log-cube transform".

```{r bpprim, fig.cap = "Postneonatal survival is Weibull vs. Nelson-Aalen.", echo = TRUE}
fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
plot(fit, fn = "cum", col = "blue")
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
```

Not very good fit. However, we need a *left-truncated Weibull distribution*, that is, assuming that
*exogeneous infant mortality* is Weibull distributed. Look here (Figure \@ref(fig:ltruw)). A perfect fit, 
just like the one under the $g$- transform. This is the way to go, *more will come tomorrow*!

```{r ltruw, fig.cap = "Postneonatal survival is truncated Weibull vs. Nelson-Aalen.", echo = TRUE}
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
plot(fit, fn = "cum", col = "blue")
xx <- hazards(fit.w)
lines(xx$x, xx$y - xx$y[1], col = "red", lty = 2)
```

@@

Remains to be written.

>>>>

Figure \@ref(fig:map) shows the study area within Sweden, with the weather stations 
marked. The map is taken (with permission) from the paper by @junkka.

```{r map, fig.height = 5, fig.cap = "Umeå, Skellefteå (orange) and its inland (green)."}
include_graphics("myfigs/map2.png")
```

@@

```{r neoanalw, results = 'asis'}
##library(survival)
winter <- neo[neo$season %in% c("winter", "spring"), ]
winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = winter)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanalw", keep = NULL)
```

@@

```{r neoanals, results = 'asis'}
##library(survival)
summer <- neo[neo$season %in% c("summer", "fall"), ]
summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = summer)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanals", keep = NULL)
```

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = winter, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
```

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = summer, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
```

@@

A modern version of the Bourgeois-Pichat biometric modelling [@bp51a; @bp51b] is given here,
*without* the "log-cube transform". A first try is fit a *Weibull* distribution to the time
to death for an infant alive at age 28 /365 years (28 days), that is, time is measured from
age 28 days until death or time reaches 365 - 28 days, when follow-up is censored.
In **R** language, with the package *eha* [@eha], we try

```
fit.w <- phreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pneo)
```

where `phreg` performs the Weibull regression (no covariates) and `coxreg` the Cox regression. In 
Figure \@ref(fig:bpprim) the result is shown in the form of the two cumulative hazards estimates.

```{r bpprim, fig.cap = "Postneonatal survival, Weibull vs. Nelson-Aalen."}
fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
plot(fit, fn = "cum", col = "blue")
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
```

Not a very good fit. However, we need a *left-truncated Weibull distribution*, that is, assuming that
*exogeneous infant mortality* is Weibull distributed.

```
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
```
Note the difference: No subtraction of the 28 days from the time variables, which implies that the time origin
is *birth* and *not* the 28th day of life. Thus our data are *left truncated* at 28 days, and the 
Weibull attempt estimates neonatal exogeneous mortality through *extrapolation*. The Cox regression approach
is a non-parametric procedure, and thus not capable of any extrapolated estimation, so neonatal
mortality will simply be taken as zero.

```{r ltruw, fig.cap = "Postneonatal survival, truncated Weibull vs. Nelson-Aalen."}
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
oldpar <- par(mfrow = c(1, 2))
##check.dist(fit, fit.w)
cc <- hazards(fit)
plot(cc[[1]][, 1], cumsum(cc[[1]][, 2]), type = "l", col = "blue", ylim = c(0, 0.07), axes = FALSE,
     xlab = "Age", ylab = "Cumulative hazards")
axis(1, at = c(28 / 365, 1), labels = c("28/365", "1"))
axis(2, las = 1)
box()
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
plot(cc[[1]][, 1], cumsum(cc[[1]][, 2]), type = "l", col = "blue", ylim = c(0, 0.07), axes = FALSE,
     xlab = "Age", ylab = "Cumulative hazards")
##plot(fit, fn = "cum", col = "blue", ylim = c(0, 0.07), axes = FALSE)
axis(1, at = c(28 / 365, 1), labels = c("28/365", "1"))
axis(2, las = 1)
box()
lines(xx$x, xx$y - xx$y[1], col = "red", lty = 2)
par(oldpar)
```

 Look here (Figure \@ref(fig:ltruw))! A perfect fit, 
just like the one under the $g$- transform. But the Weibull curve
does not start at zero (left hand panel), however it is parallell 
to the Nelson-Aalen estimator, and the right hand panel shows what happens if we subtract 
the start value from the Weibull curve and redraw.

## Separate exogeneous and endogeneous Infant mortality

So, finally, we show the full procedure how to perform the "biometric analysis" *without*
the log-cube transform. It simply consists of two steps:

1.    Run a Cox regression on the full data set.
2.    Run a *left truncated* (at 28 days) Weibull regression.
3.    Plot the two cumulative hazards functions in the same graph and read
      their crossings of the vertical line at age = 1 year (365 days). Figure \@ref(fig:crossings).

```{r crossings, fig.cap = "Biometric analysis.", fig.height = 4}
fit.c <- coxreg(Surv(enter, exit, event) ~ 1, data = infdat)
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
cc <- hazards(fit.c)
plot(c(0, cc[[1]][, 1]), c(0, cumsum(cc[[1]][, 2])), type = "l", col = "blue", axes = FALSE,
     xlab = "Age (days)", ylab = "Cumulative hazards", ylim = c(0, 0.08))
axis(1, at = c(0, 28 / 365, 1), labels = c(0, 28, 365))
axis(2, las = 1, at = round(c(0, max(cumsum(cc[[1]][, 2])), max(xx$y)), 3))
##abline(v = 1)
box()
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
abline(h = 0, v = 0)
abline(v = 1, lty = 3)
abline(v = 28 / 365, lty = 3)
abline(h = c(max(cumsum(cc[[1]][, 2])), max(xx$y)), lty = 3)
```

So, total mortality is 0.076, exogeneous mortality is 0.063, and endogeneous mortality
is the difference, 0.013, or 17 per cent of the total mortality.

Now, try the original "biometric" approach and compare *(Exercise)*.

@@

Remains to be written, especially the discussion about temperature and mortality.
